\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-04 NAR_DRAFT.tex
lunch changes

\section{DISCUSSION}

\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.

The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010} .
These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.

\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
These high-probability variants are generated from the mutation profile of the tumor, and measured with very small error.
This data allows us to reliably estimate the distribution of substitution types.
Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.

The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors.
Deep sequencing is useful \cite{Zehir2017}.

@@

>>>>

\section{DISCUSSION}
We hope that the availability of this algorithm will have a positive effect on the cost/benefit ratio of deeper sequencing in both research and clinical applications
One size does not fit all with respect to the evidence threshold. Users are interested in true and false positive fractions, and users of different types will have different trade-offs in terms of sensitivity and specificty
Clinical studies often threshold at a vaf, rather than using the evidence threshold of the caller. 272X 5\% cutoff \cite{Parsons2016}, why go so deep?
One problem with this is that vaf and associated evidence is dependent on local depth. This is no different than just having a minimum number of variant reads. Why have a model?
The benefit derived from \batcave varies with both the concentration of the tumor mutation profile and the depth of sequencing.
Deeper sequencing leads to more benefit from the algorithm because there are simply more low frequency, and thus more ambiguous, variants accessible in deeply sequenced tumors.
Deep sequencing is useful \cite{Zehir2017}.
\bkmcomment{the above should be some version of, variant calling is important, deeper is better, we are taking a step in that direction}

\batcave is a post-call variant evalutation tool that uses MuTect output and the tumor-specific mutation profile to sharpen the posterior probability assigned to SNVs.
We demonstrated here that \batcave improves classification accuracy in simulated data, with the magnitude of the improvement dependent on the sequencing depth and the concentration of the mutation profile.
\bkmcomment{deeper summary of results, and some meaning Somithing about calibration dummy}.

The \batcave algorithm focuses on the tri-nucleotide context and substitution type, which has been shown to have a large effect on site-specific mutation rates \cite{Martincorena2015,Hollstein2017}.
There are, however, many other types of \textit{genomic context} which can effect local mutation rates including replication timing, expression level, and chromosome organization \cite{Buisson2019,Schuster-Bockler2012,Pleasance2010}.
\bkmcomment{what to say here? Are we acknowledging a weakness, or just observing there are other types?}
These could be very useful in computing a local or regional mutation rate, as opposed to the genome-wide but tumor-specific mutation rate used by \batcave.

\bkmcomment{summarize run-time here}
We have chosen to implement \batcave as a post-calling algorithm for exclusive use with MuTect because MuTect presents a simple model and a well-tested set of heuristic filters.
However, the model is very general, and in principle there is no barrier to using the method with other variant callers.
For instance Strelka2, another widely used variant caller, computes a joint posterior probability over tumor and normal genotypes where the prior probability that the tumor carries a somatic mutation at a given site is set to a constant \cite{Kim2018}.
Modifying the algorithm to work with the Strelka statistical model would involve replacing the constant prior with a tumor and site-specific prior for each potential variant, exactly as we have done with MuTect.
The majority of the computational overhead in \batcave is associated with extracting the tri-nucleotide context for each potential variant site from the reference genome.
Since most variant callers are already walking the reference genome as part of the calling process, extracting the tri-nucleotide context simultaneously would dramatically reduce the computational overhead associated with implementing \batcave.

\bkmcomment{can I work a citation in here? using data sets from the same data generating distribution as a prior?}
\batcave approaches the set of variant calls returned from MuTect as two distinct sets of data.
The first is the set of mutations that pass all heuristics filters, and  have a very large posterior probability of being present given the observed read data.
This data allows us to reliably estimate the distribution of substitution types, i.e. the mutation profile.
Variants with low probability given the read data are noisy measurements of samples from the same data generating distribution.
We use the estimated distribution of substitution types to adjust the prior probability of mutation for these lower probability variants.

@@

\section{Software Availability}
github for source code
zenodo for figures, simulations
zenodo for package at time of bioarxiv
